---
title: "615_Assignment3_Honey"
author: "Dave Anderson, Sky Liu, Tingrui Huang, Xiang Xu"
date: "October 3, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}
#Loading data files
Honey <- read_csv("Honey.csv")
Deadout <- read_csv("Deadout.csv")
Price_per_lb <- read_csv("Price per lb.csv")
Production_per_Colony <- read_csv("Production per Colony.csv")
Honey_value_annual <- read_csv('Honey_value.csv')
CPI <- read_csv('1987_2017CPI.csv')
honey_loss_dt <- read.csv('Honey_Loss_6_States.csv')
```

```{r}
#sort each data set to variables we want. 
Honey <- dplyr::select(Honey,Year,State,Value)
Deadout <- dplyr::select(Deadout,Year,Period,State,Value) 
Price_per_lb <- dplyr::select(Price_per_lb,Year,State,Value)
Production_per_Colony <- dplyr::select(Production_per_Colony,Year,State,Value)
Honey_value_annual <- dplyr::select(Honey_value_annual,Year,Value)



#Filter out totals, group each variable by state, average values from each year
Production <- filter(Honey, State != "US TOTAL") %>% group_by(State) %>% summarise(Average_production = mean(Value)/2000)
Loss <- filter(Deadout, State != "US TOTAL") %>% group_by(State) %>% summarise(Average_loss = mean(Value))
Price <- filter(Price_per_lb, State != "US TOTAL") %>% group_by(State) %>% summarise(Average_price = mean(Value))
Colony_production <- filter(Production_per_Colony, State != "US TOTAL") %>% group_by(State) %>% summarise(Average_per_colony = mean(Value))

#Combine into one set. Add new variables to show number of colonies and loss/colony
Honey_by_State <- full_join(Production, Loss, by = "State")
Honey_by_State <- full_join(Honey_by_State, Price, by = "State")
Honey_by_State <- full_join(Honey_by_State, Colony_production, by = "State") %>% 
  mutate(Colonies = Average_production*2000/Average_per_colony) %>% 
  mutate(Loss_per_colony = Average_loss/Colonies) 

# pick 6 top states with highest production and complete data
Honey_State <- Honey %>%  group_by(State) 
unique(Honey$State)
Honey_sixstate <- Honey_State %>%
  filter(State %in% c("CALIFORNIA", "FLORIDA" ,"SOUTH DAKOTA", "NORTH DAKOTA","MONTANA", "MINNESOTA")) %>%
  arrange(State, Year)

#Honey lost in 6 states
#Sum by year (Since we only have the data in 1st and 2nd quarter in 2018, we will exclude the data in 2018)
honey_loss_dt$Value <- as.numeric(gsub(",","",honey_loss_dt$Value))
honey_2017 <- honey_loss_dt %>% select(Year,State,Value) %>% filter(Year==2017) %>% group_by(Year,State) %>% summarise(total=sum(Value))
honey_2016 <- honey_loss_dt %>% select(Year,State,Value) %>% filter(Year==2016) %>% group_by(Year,State) %>% summarise(total=sum(Value))
honey_2015 <- honey_loss_dt %>% select(Year,State,Value) %>% filter(Year==2015) %>% group_by(Year,State) %>% summarise(total=sum(Value))
# Total loss from 2015-2017
honey_total <- rbind(honey_2017,honey_2016,honey_2015)


#Adjust the annual honey value (price received) by 1987 inflation rate.
baseCPI <- rep(113.6, 21)
adjusted_Price <- as.data.frame(Honey_value_annual$Value * (CPI$CPI / baseCPI))
Honey_value_annual <- cbind(Honey_value_annual,adjusted_Price)
names(Honey_value_annual) <- c('Year','Value','adjValue')
#Add annual productivity
Annual_production <- filter(Honey, State != "US TOTAL") %>% group_by(Year)  %>% summarise(Average_production = mean(Value)/2000)  
Annual_production <- arrange(Annual_production, desc(Year))
Honey_value_annual <- cbind(Honey_value_annual,Annual_production$Average_production)
names(Honey_value_annual) <- c('Year','Value','adjValue','annualProd')

```





```{r}
#Change of Honey annual value and productivity from 1987-2017
ggplot(data = Honey_value_annual, aes(x = Year)) + 
  geom_line(aes(y = adjValue, colour = "adjValue"))+
  geom_line(aes(y = annualProd, colour = "annualProd"))  +
  ggtitle("Author: Sky Liu \n Change of Honey annual value and productivity from 1987-2017")+
  scale_y_continuous(sec.axis = sec_axis(~.*5, name = "Annual productivity [%]"))+
  scale_colour_manual(values = c("blue", "red"))+
  labs(y = "Adjusted Value [$]",
                x = "Year")

#Scatterplot to show relationship between productivity and loss by state.
#Colored to show overall high-producing states.
ggplot(data = Honey_by_State, mapping = aes(Average_per_colony,Loss_per_colony))+
  geom_point(aes(color = Average_production),size = 3)+
  ggtitle("Author: Dave Anderson \n Relationship b/t productivity and loss by state")


#Total Production by state, largest six states indicated in red. 
ggplot(data = Honey_by_State, mapping = aes(State, Average_production, fill=ifelse(Average_production > 5000,"A", "B")))+
  geom_col()+
  scale_fill_manual(guide=FALSE, values=c("red", "black"))+
  theme(axis.text.x = element_text(color = ifelse(Honey_by_State$Average_production > 5000, "red", "black"), angle = 75, hjust = 1))+
  ggtitle("Author: Dave Anderson \n Production by state")


#Display of prices per lb. by state. Top six producing states still in red to show their low prices. 
ggplot(data = Honey_by_State, mapping = aes(State ,Average_price, fill = ifelse(Average_production > 5000, "A", "B")))+
  geom_col()+
  scale_fill_manual(guide=FALSE, values=c("red", "black"))+
  theme(axis.text.x = element_text(color = ifelse(Honey_by_State$Average_production > 5000, "red", "black"), angle = 75, hjust = 1))+
  ggtitle("Author: Dave Anderson \n Prices per lb. by state")

#trend of annual value of 6 top productivity states 
ggplot(Honey_sixstate, aes(x=Year, y=Value ,color = State)) +
  geom_line(size = 1) +
  ggtitle("Author: Xiang XU \nValue year trend by state")

# annual honey lost trend of 6 top productivity states from 2015-2017
ggplot(honey_total, aes(x=Year, y=total, color=State))+geom_line()+
  ggtitle("Author: Tingrui Huang \n Annual honey lost trend of 6 top productivity states from 2015-2017")

#Display of production vs. loss per colony with out top states labeled. 
ggplot(data = Honey_by_State, mapping = aes(Average_production,Loss_per_colony, label = State))+
  geom_point() + geom_text(aes(label = ifelse(Average_production > 5000,as.character(State),'')),vjust = 1.5,hjust = 1)+
   ggtitle("Author: Dave Anderson \n Production vs. loss per colony with out top states labeled")

#Display of colonies lost by year and peiod. 
loss_by_year <- Deadout %>% filter(State == "US TOTAL") %>% group_by(Year)
ggplot(data = loss_by_year)+
  geom_col(mapping = aes(x = Year, y = Value, fill = Period))+
  ggtitle("Author: Dave Anderson \n Colonies lost by year and peiod")



```

#It is interesting to see two of our top producers from, big, southern,
#costal states while the other 4 are from the midwest. 

#Looking at prices, we see that the big producers are also among cheapest states.

#The two large states have high rates of deadout colonies. The 4 big producers from the midwest have low loss rates. 

*We pick six states, which have 31 years data, CALIFORNIA, FLORIDA , SOUTH DAKOTA, NORTH DAKOTA, MONTANA and MINNESOTA. Accordng to the graph above, we can see the trend about annual value of each state for 31 years.*

*Clearly, NORTH DAKOTA is the only state among these six states, whose yearly value has a upward trend in last 31 years. The other five states decline volatility in year-values.*

#As we can see in the graph, California has the highest loss all the time, but there was a decrease
#from 2016 to 2017. Florida and North Dakota has very close loss in 2015 and the begining of 2016, 
#however, North Dakota has decreased its loss by 35% from 2016 to 2017 and it now has lower loss
#than Florida. We won't be suprised if there is an decrease in the Price of honey and increase in the
#Production of honey in North Dakota in the year of 2016-2017. The other three states have steady 
#low loss comparing with California, Florida and North Dakota.